---
title: "Final Write Up"
author: 
  - Ava Wessel
  - Mia Hanson
output:
  pdf_document: default
  html_document:
    df_print: paged
date: "2024-12-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MuMIn)
library(lubridate)
library(tidyverse)
library(sp)
library(sf)
library(raster)
library(prism)
library(terra)
library(lme4)
library(lmerTest)
library(car)
library(ggplot2)
library(ggeffects)
library(ggpubr)

setwd("~/Bee-Urbanization-BI410")

complete_bee_data <- read_rds("Data/final_bee_dataset")

```

Note to reader: We had to combine four different R markdown scripts, so the addition of saving our models/datasets and rereading in the datasets was set up prior to combination. I guess it is still a good habit to have though. Because of this, the code from the original scripts might differ slightly to get this file to properly knit. The overall function of this markdown is the same. 

Effect of Urbanization and Temperature on Bombus spp. Emergence

I. Introduction: 

Bumble bees (Bombus spp.) and other bee species play an important ecological role as pollinators for various wildflowers and agricultural crops (Ahrné et al. 2009; McCabe et al. 2022). Despite their ecological importance, pollinators face numerous threats from habitat loss, pesticide exposure, climate change, and disease, with many bee populations declining worldwide (Goulson et al. 2008). Habitat loss and fragmentation caused by urbanization emerges as a growing concern due to its rapid expansion and transformation of landscapes that were once natural habitats into fragmented urban environments. Urbanization can have a large effect on Bombus spp. flower visitation rates and pollination success through changes in foraging behavior (Glaum et al. 2017; Harrison & Winfree 2015). Urban heat islands, or the increased temperatures often observed in cities due to concrete and reduced green spaces, further complicate the challenges faced by pollinators (Harrison & Winfree 2015). Together, these pressures can lead to declines in Oregon bee populations, leading to cascading effects impacting the numerous ecosystem services offered. Understanding how urbanization impacts emergence patterns of Bombus spp. is crucial, as these factors are closely tied to their ability to effectively pollinate and maintain stable populations.

Recent research has found that climate change, often resulting from urbanization, is altering phenology, with spring events generally occurring earlier (Stemkovski et al. 2020). For Bombus spp., increased (spring) temperatures have been shown to directly lead to advanced phenology, prompting earlier emergence and activity periods (Prestele et al. 2021). Harrison & Winfree (2015) observed that urban habitats support a greater abundance of late-season bees compared to forest habitats, likely due to the presence of late-blooming plants. They also suggest that urban warming may act as an "ecological filter," reducing species diversity and resulting in a more homogenized bee community in cities. Similarly, Blasi et al. (2023) found that over the past 20 years, a citizen-science dataset has shown that Bombus spp. emergence across Sweden has advanced significantly, though responses varied based on species traits. As in Harrison & Winfree’s findings, warmer temperatures substantially advanced the flight period for early-emerging species, while late-emerging species exhibited no consistent trend. This variability in response among Bombus spp. may contribute to shifts in community composition, favoring species that can better synchronize their life cycles with the altered urban floral availability. Little is known about how these phenological changes impact pollination success across diverse urban ecosystems, especially in areas like Oregon where agricultural productivity and biodiversity heavily depend on stable bee populations.

Understanding these emergence patterns and phenological shifts is crucial for pollinator diversity conservation efforts. Using the Oregon Bee Atlas (OBA), census population block data, and monthly temperature reading from the PRISM group, we are able to develop a more comprehensive analysis of how urbanization and temperature variability affect Bombus spp. Li et al. (2020) demonstrated that human population density can serve as a reliable indicator of urbanization, allowing us to use census data as a proxy for assessing the intensity of urban development and its impacts on bee populations. By combining monthly temperature data and population density, we are able to assess the response of Bombus spp. Across various levels of urbanization, allowing us to identify species-specific emergence patterns and population trends among different bee species.

Our research question is: How does peak emergence for four Bombus species, vosnesenskii, melanopygus, mixtus, and griseocollis, vary along climatic and urbanization gradients? Our first hypothesis is that bees in more heavily urbanized areas will have earlier emergence dates. Our second hypothesis is that areas that have higher mean temperatures will show earlier emergence times. Our third hypothesis is that earlier emerging species will be more sensitive to warming temperature than later emerging species, specifically we will see a more significant change in emergence days per °C of temperature change. 


II. Data/ Data Wrangling: 

The data that we be combined is the OBA dataset and merged it with US Census data to get whether those bee observations were in a more rural or urban area. We read in the US Census block data as a shapefile data type. Then we wrangled the population column into 4 even categories using the quantile function. The 4 categories were rural, subrural, urban, and suburban. We then read in the PRISM temperature data, originally as raster data, merging each of the bee observations with temperature, joining on latitude and longitude in order to see what the average temperature is in the area the bee was recorded. With the PRISM temperature data we calculated average spring temperatures, finding the mean of the average temperature from months February-June. For data exploration we made a couple tables filtering the species we were considering focusing on for our analysis, in the Data_Exploration.Rmd file and we took a look at any weird values like NAs in our dataframes using tables in the Data_Cleaning.Rmd file.

```{r}
#Loading in OBA
bee_data <- read.csv("OBA_2018-2023_combined_dataset_working.csv")

# want to just keep our species from the main dataframe
filtered_bee_data <- bee_data[bee_data$Species %in% c("vosnesenskii", "melanopygus", "mixtus", "griseocollis"), ]

# want to keep Dec..Lat. and Dec..Long. and Collection.Date and Collection.Day.1
filtered_bee_data <- filtered_bee_data %>%
    dplyr::select(Dec..Lat., Dec..Long., Collection.Date, Species, Year.1)
 
head(filtered_bee_data)

# removing these na rows 
filtered_bee_data <- na.omit(filtered_bee_data)
```

# creating a day of year column
```{r}
# logic behind the day of year: January 1st is day 1, jan 2nd is day 2 etc 

  # using lubridate

calc_day_of_year <- function(date){
  
  # converting the string object to a Date object
  # parses mm/dd/yyyy
  date <- mdy(date) 
  
  # get day of year
  day_of_year <- yday(date)
  
  return (day_of_year)
  
}

filtered_bee_data <- mutate(filtered_bee_data, DOY = calc_day_of_year(Collection.Date))

#Creating winter month, day, and year for PRISM extraction 
month <- seq(as.Date("2020-01-01"), 
             as.Date("2020-12-01"), 
             by = "1 month")
month_label <- lubridate::month(month, label = TRUE)

winter.months <- seq(as.Date("2020-09-01"), 
                     as.Date("2021-08-01"), 
                     by = "1 month")
winter.month_label <- lubridate::month(winter.months, label = TRUE)

filtered_bee_data <- filtered_bee_data %>%
  mutate(Collection.Date=mdy(Collection.Date)
         ,year=as.numeric(year(Collection.Date))
         ,month=month(Collection.Date)
         ,doy = yday(Collection.Date)
  ) %>%
  arrange(year,doy) 

filtered_bee_data <- filtered_bee_data %>% rename(latitude = Dec..Lat., longitude = Dec..Long.)

filtered_bee_data <- mutate(filtered_bee_data,
                 winter.year = ifelse(month>8, year+1, year)
                 ,winter.month = ifelse(month>8, month-8, month+4)
                 ,winter.day = ifelse(month>8, doy-243, doy+120)
)


filtered_bee_data <- na.omit(filtered_bee_data)

dim(filtered_bee_data)

```


# Adding in PRISM climate data:

Note: I could not for the life of me figure out how to bypass this knitting error without completely changing the code. I check to see if they were all numeric and they were, not really sure the solution. To see working code on the original markdown file, see Data_Cleaning.rmd. I included the code up until it stopped working and the execution halt error occurs. I included the error below: 
Error in `FUN()`:
! only defined on a data frame with all numeric-alike variables
Backtrace:
  1. ... %>% print(bee_long)
  4. pillar:::print.tbl(., bee_long)
  5. pillar:::print_tbl(...)
  8. pillar:::format.tbl(...)
  9. pillar:::format_tbl(...)
     ...
 17. pillar:::ctl_new_pillar.tbl(controller, vector(), width, title = title)
 20. pillar:::max0(width)
 21. base::Summary.data.frame(`<tibble[,21]>`, na.rm = FALSE)
 22. base::lapply(...)
 23. base (local) FUN(X[[i]], ...)

Quitting from lines 130-202 [unnamed-chunk-3] (Final_Writeup.Rmd)
Execution halted

```{r}
#Adding PRISM Data and calculating values for 'winter year' so that we take into account the months prior to the spring season that would be affecting what we see.

prism_set_dl_dir("prism_data")

# Had to comment out as it was printing all output in the pdf
#get_prism_monthlys(
# type = "tmean", 
#  year=2016:2021,
#  mon=1:12,
#   keepZip = FALSE
# )

sub.tmean <- prism_archive_subset("tmean", "monthly", mon = 1:12)

RS <- pd_stack(sub.tmean) ##raster file  

bee.test <- filtered_bee_data
bee.spdf <-   SpatialPointsDataFrame(coords=bee.test[,c('longitude','latitude')], 
                                    data=bee.test, proj4string = CRS("+proj=longlat +ellps=WGS84 +no_defs"))

bee.clim <- extract(RS, bee.spdf,  fun=mean, na.rm=TRUE, sp=TRUE) 
bee.clim <- as.data.frame(bee.clim)

#...(rest of code in Data_Cleaning.rmd)

# Read-in from original script due to knitting error 
bee_data_wTemp <- read_rds("Data/bee_data_wTemp")


```

Census Block Data
```{r}
# Adding in Census Block Data

census_shp <- st_read("Data/CensusBlockGroups/CensusBlockGroups.shp")

census_sf <- census_shp %>% 
  st_transform(crs = "EPSG:4326") %>% 
  rename(latitude = INTPTLAT, longitude = INTPTLON) %>% 
  dplyr::select(latitude, longitude, geometry, POP20, POPDENS20, COUNTY)
  

bee_sf <- st_as_sf(bee_data_wTemp,
                        coords = c("longitude", "latitude"),
                        crs = crs("EPSG:4326"))
census_sf <- st_cast(census_sf, "POLYGON")

bee_with_census <- st_join(bee_sf, census_sf)
dim(bee_with_census)

saveRDS(bee_with_census, "Data/final_bee_dataset")


```

III. Analysis / Hypothesis Testing:

To test our hypothesis we used a linear mixed model, and we completed our analysis in the file Bee_Analysis.Rmd. We predicted the day of year of peak emergence for our bee species. We ran the model on all of the species together and on each species separately to see the specific effects of temperature and urbanization on each species. Our formula was: DOY ~ 1 + urbanizationLevel + springTemp + pop + (1 | species). So our fixed effects to predict the day of year were urbanization level, average spring temperature, and population. And our random effect was species to account for variation between species. We ran this model using the “lmer” function. We also used backward selection and the Variance Inflation Factor to make sure that the formula we were using was the most effective at predicting day of year and that we didn’t have any concerning covariance in our model. The output of our model found significance (p < 0.05) in each of our predicting effects, showing that when compared to rural, there was a significant difference in emergence time. Specifically the effect of urbanization resulted in an earlier emergence of about 15.8 days per °C compared to rural areas. Warmer spring temperatures resulted in an earlier emergence day by about 8.5 days per °C. We also found that Bombus spp. that on average have an earlier peak emergence were more sensitive to changes in temperature, going along a gradient where the later peak emergence species were found to be gradually less sensitive to temperature changes.  


Adding urbanization levels 
```{r}
complete_bee_data <- read_rds("Data/final_bee_dataset")

# function to add urbanization levels, rural, subrural, suburban, and urban
  # split into 4 even categories using the quantile function
assign_urbanization_level <- function(data) {
  quantiles <- quantile(data$POP20, probs = seq(0 , 1, 0.25), na.rm = TRUE)
  
  data <- data %>% 
    mutate(urbanization_level = cut(POP20,
                                    breaks = quantiles,
                                    labels = c("rural", "subrural", "suburban", "urban"),
                                    include.lowest = TRUE))
  
  return(data)
}


complete_bee_data <- assign_urbanization_level(complete_bee_data)
saveRDS(complete_bee_data, "~/Bee-Urbanization-BI410/Data/complete_bee_data")

```

Finalizing the columns and data that we want to use for our model and running the model
```{r model}

# creating the column spring temp, averaging temperatures from Feb to June
complete_bee_data <- complete_bee_data %>%
  rowwise() %>%
  mutate(springTemp = mean(c_across(c("tmean_6", "tmean_7", 
                                      "tmean_8", "tmean_9", "tmean_10")), na.rm = TRUE)) %>%
  ungroup()

# creating a dataframe with all of our variables that we want to use in the model
dat <- data.frame(
  DOY = complete_bee_data$DOY,
  springTemp = scale(complete_bee_data$springTemp),
  urbanizationLevel = complete_bee_data$urbanization_level,
  species = complete_bee_data$Species,
  pop = scale(complete_bee_data$POP20)
)

sum(is.na(dat))
dat$springTemp <- as.numeric(dat$springTemp)


# Model 
formula <- DOY ~ 1 + urbanizationLevel + springTemp + pop + (1 | species)

model1 <- lmer(formula, data = dat)

summary(model1)
step(model1)

vif(model1)

saveRDS(model1, "~/Bee-Urbanization-BI410/model1")

# calculating R^2 values
r2 <- r.squaredGLMM(model1)
r2

```

## Applying the model to each of our 4 species
```{r}
# MOdel applied to each species 

select_species <- c("vosnesenskii", "melanopygus", "mixtus", "griseocollis")

model_results <- list()
r2_val <- list()

for (i in seq_along(select_species)) {
  
  formula_for_species <- DOY ~ 1 + urbanization_level + scale(springTemp) + scale(POP20) + (1 | COUNTY)
  
  single_species <- na.omit(complete_bee_data[complete_bee_data$Species == select_species[i], ])
  
  species_model <- lmer(formula_for_species, data = single_species)
  model_results[[select_species[i]]] <- species_model
  r2 <- r.squaredGLMM(species_model)
 r2_val[i] <- r2[2]
}

model_summaries <- lapply(model_results, summary)

model_summaries[1]
model_summaries[2]
model_summaries[3]
model_summaries[4]

r2_val

saveRDS(model_summaries, "~/Bee-Urbanization-BI410/species_model_summaries")

```


IV. Visualizations: 

These first two figures were made to summarize the raw data in two different ways. 

Bombus Observations across Oregon:
```{r Bombus observations across oregon}

# making our data into a shapefile for plotting
bombus_shp <- st_as_sf(complete_bee_data,
                                   coords = c("Dec..Long.", "Dec..Lat."),
                                   crs = st_crs("EPSG:4326"), na.fail = FALSE	
)

# Calculating the same average spring temperatures that we used for the model 
complete_bee_data <- bombus_shp %>%
  rowwise() %>%
  mutate(springTemp = mean(c_across(c("tmean_6", "tmean_7", 
                                      "tmean_8", "tmean_9", "tmean_10")), na.rm = TRUE)) %>%
  ungroup()

# downloading in the Oregon base map 
or <- map_data("county", "oregon")%>% 
  dplyr::select(lon = long, lat, group, id = subregion)

# first need to create the extent from the or dataframe
min_long <- min(or$lon)
max_long <- max(or$lon)
min_lat <- min(or$lat)
max_lat <- max(or$lat)

or_extent <- c(xmin = min_long, xmax = max_long, ymin = min_lat, ymax = max_lat)

# create bounding box
bbox <- st_bbox(or_extent, crs = st_crs(bombus_shp))
# concert bbox to sf polygon
bbox_poly <- st_as_sfc(bbox)
# crop shapefile using bounding box
cropped_bombus_shp <- st_intersection(bombus_shp, bbox_poly)


# plotting!
ggplot() +
  geom_polygon(data=or, aes(x=lon, y=lat, group=group)) +
  geom_sf(data = cropped_bombus_shp, aes(color = Species)) +
  scale_color_viridis_d(name = "Species") + 
  ggtitle("Bombus Species Observations Across Oregon") + 
  labs(x = "longitude", y = "latitude") + 
  scale_color_manual(
    values = c("red", "white", "forestgreen", "violet"), # Custom colors for each species
    name = "Species"
  ) 
```
Figure 1: Map showing the average spring temperatures across Oregon, each point is a bee observation.


Census Blocks Plot
```{r Census Blocks Plot}

census_shp <- st_read("Data/CensusBlockGroups/CensusBlockGroups.shp")
census_csv <- read.csv("CensusBlockGroups.csv")

#ggplot() +
#  geom_polygon(data=or, aes(x=lon, y=lat, group=group), fill = "tan")

census_shp <- census_shp %>%
  mutate(pop_bin = cut(POP20, breaks = c(0, 3000, 4000,5000, 6000, Inf),
                       labels = c("0-3k", "3k-4k","4k-5k", "5k-6k", "6k+"),
                       include.lowest = TRUE))

ggplot() +
  geom_sf(data = census_shp, aes(fill = POP20)) + 
  scale_fill_gradient(low="beige", high = "brown", name = "Population by Census Block") + 
  
  geom_sf(data = complete_bee_data, aes(color = springTemp)) +
  scale_color_gradient(low="blue", high="red", name = "Avg Spring Temp °C (Feb-June)") + ggtitle("All Bee Locations with Census Blocks and Spring Temperature Info") + theme(plot.title = element_text(hjust = 0.3, vjust = 3)) + 
  
  xlab("lon") + ylab("lat") 

```
Figure 2: Map with average spring temperatures and population by Census Block across Oregon, each point is a bee observation.


The three figures below showcases the predicted linear trend for spring temperature, urbanization, and species specific response. 

Creating Temperature variables and checking linear model. (model1):
```{r}
complete_bee_data <- complete_bee_data %>%
  rowwise() %>%
  mutate(springTemp = mean(c_across(c("tmean_6", "tmean_7", 
                                      "tmean_8", "tmean_9", "tmean_10")), na.rm = TRUE)) %>%
  ungroup()

anova(model1)

summary(model1)
```


Spring Temperature vs DOY Visualization:
```{r}

tempDOY <- ggpredict(model1, terms = c("springTemp"))


tempDOY_plot <- ggplot(tempDOY, aes(x = x, y = predicted)) +
  geom_point(data = complete_bee_data, 
             aes(x = scale(springTemp), y = DOY, color = "Observed Data"), 
             alpha = 0.5) +
  geom_line(aes(color = "Model Prediction"), linewidth = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  scale_color_manual(values = c("Observed Data" = "green3", "Model Prediction" = "darkgreen")) +
  labs(x = "Scaled Spring Temperature", y = "DOY", color = "Legend", fill = "Legend",
       title = "                  Effect of Temperature on Day of Emergence") +
  theme_minimal()

#calculating slope 
slope_tempDOY <- diff(tempDOY$predicted) / diff(tempDOY$x)
print(slope_tempDOY)


tempDOY_plot

```
Figure 3: Average peak Bombus spp. emerge earlier with increasing spring temperatures. Green line represents model prediction with 95% confidence interval. Green dots represent observed data. 


Urbanization Level vs DOY Visualization:
```{r}

urbanizationDOY <- ggpredict(model1, terms = c("urbanizationLevel"))

urbanizationDOY_plot <- ggplot(urbanizationDOY, aes(x = x, y = predicted)) +
  geom_line(aes(group = 1), size = 1) +  
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), 
                width = 0.2, size = 0.8, color = "darkgrey") +
  geom_point(size = 3, color = "forestgreen") +
  labs(x = "Urbanization Level", 
       y = "Predicted DOY",
       title = "                            Effect of Urbanization on Day of Emergence") +
  theme_minimal() 


urbanizationDOY_plot

```
Figure 4: Dense urban areas lead to an earlier average peak emergence when compared to rural areas. Green dots represent slopes for each urbanization level. Gray bars represent 95% confidence interval. 

Species Specific Figure:

```{r}

results <- data.frame(
  species = character(),
  slope = numeric(),
  conf.low = numeric(),
  conf.high = numeric(),
  stringsAsFactors = FALSE
)

for (species_name in names(model_summaries)) {
  coefficients <- model_summaries[[species_name]][["coefficients"]]
  
  if ("scale(springTemp)" %in% rownames(coefficients)) {
    slope <- coefficients["scale(springTemp)", "Estimate"]
    std_error <- coefficients["scale(springTemp)", "Std. Error"]
    df <- coefficients["scale(springTemp)", "df"]
    
    t_critical <- qt(0.975, df = df)
    conf.low <- slope - t_critical * std_error
    conf.high <- slope + t_critical * std_error
    
    results <- rbind(
      results,
      data.frame(
        species = species_name,
        slope = slope,
        conf.low = conf.low,
        conf.high = conf.high,
        stringsAsFactors = FALSE
      )
    )
  } 
}

ggplot(results, aes(x = reorder(species, slope), y = slope)) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2, color = "gray50") +
  geom_point(size = 3, color = "forestgreen") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", size = 0.8) +
  labs(
    y = "Emergence Sensitivity (days/°C)",
    x = "Species",
    title = "                  Species-Specific Emergence Sensitivity to Temperature"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1) 
  ) 

```
Figure 5: Melanopygus saw the greatest emergence sensitivity, with mixtus, vosnesenskii, and griseocollis following. Green dots represent slopes for each species. Black bars represent 95% confidence intervals for each estimate. Red dashed line at 0 represent no emergence sensitivity to temperature. 


Summary stats for species specific response comparison:

Mean DOY and total number of observations for all species. 
```{r}
doy_avg <- complete_bee_data %>% 
  group_by(Species) %>% 
  summarise(doy_avg = mean(DOY), 
            observations = sum(n())) %>% 
  arrange(doy_avg, desc = TRUE)

doy_avg
 
```


V. Conclusion: 

We found that more urbanized areas resulted in an earlier peak emergence, by around 15.8 days per °C. This is likely caused by the Urban Heat Island Effect which causes warmer microclimates due to less green spaces and a higher concentration of concrete in more urbanized areas. The effect of temperature also causes earlier emergence dates, across all species, by around 8.5 days per °C (more specifics on model predictions can be found in the Bee_Analysis.rmd script). This lines up with our understanding of temperature as an environmental cue that regulates phenology, directly or through changes in plant phenological sensitivity. The effect of urbanization and temperature on bees might indirectly affecting them, as studies have shown that plant phenology is highly synchronized with temperature. Further research would need to be conducted assess the relationship between the plant the bee was found on, the climate, and the bees species itself. As for the species-specific phenological sensitivity, Bombus species that have an earlier average peak emergence were more sensitive to the warming changes in temperature, the reliance on temperature could make early-season species more vulnerable to climate change with premature emergence that could cause a mismatch with floral resource availability. This result aligns with our initial hypothesis as melanopygus, the earliest emerging species also is the most sensitive to temperature. As you move to later emerging Bombus species, the less sensitive to temperature they appear to be. 
	
Things that could be causing bias include the fact that urban areas are more populated and easier to access, so there might be an overestimation of the effects of urbanization due to the increased number of observations in urban areas. We also have to consider, due to our usage of quantiles, just because something was labeled urban or rural, there is a chance it might not fit the true description of what a rural or urban environment looks like. The choice of creating categorical values for urbanization allowed us to easily test the two extremes, but possibly ignored what the true defined urbanization levels are for a given observation. Additionally, there are more common species with more observations: Bombus vosnesenskii had 2,695 observations, Bombus mixtus had 1549, Bombus melanopygus had 566, and Bombus griseocollis had 306. Because of the substantially lower observations for melanopygus and grisecollis, the possibility of skewing their data within the overall model was feasible. Though we created a species specific model to account for variations in observation, this wasn't taken into account within the overall model. More specifically, we could be underestimating the effects on species like griseocollis and overestimating the effects of species like vosnesenskii. 


VI. References:

Ahrné, K., Bengtsson, J., & Elmqvist, T. (2009). Bumble Bees (Bombus spp) along a Gradient of Increasing Urbanization. PLoS ONE, 4(5), e5574. https://doi.org/10.1371/journal.pone.0005574
McCabe, L. M., Aslan, C. E., & Cobb, N. S. (2022). Decreased bee emergence along an elevation gradient: Implications for climate change revealed by a transplant experiment. Ecology, 103(2), e03598. https://doi.org/10.1002/ecy.3598
Harrison, T., & Winfree, R. (2015). Urban drivers of plant‐pollinator interactions. Functional Ecology, 29(7), 879–888. https://doi.org/10.1111/1365-2435.12486
Glaum, P., Simao, M.-C., Vaidya, C., Fitch, G., & Iulinao, B. (2017). Big city Bombus : using natural history and land-use history to find significant environmental drivers in bumble-bee declines in urban development. Royal Society Open Science, 4(5), 170156. https://doi.org/10.1098/rsos.170156
Goulson, D., Lye, G. C., & Darvill, B. (2008). Decline and Conservation of Bumble Bees. Annual Review of Entomology, 53(1), 191–208. https://doi.org/10.1146/annurev.ento.53.103106.093454
Stemkovski, M., Pearse, W. D., Griffin, S. R., Pardee, G. L., Gibbs, J., Griswold, T., Neff, J. L., Oram, R., Rightmyer, M. G., Sheffield, C. S., Wright, K., Inouye, B. D., Inouye, D. W., & Irwin, R. E. (2020). Bee phenology is predicted by climatic variation and functional traits. Ecology Letters, 23(11), 1589–1598. https://doi.org/10.1111/ele.13583
Blasi, M., Carrié, R., Fägerström, C., Svensson, E., & Persson, A. S. (2023). Historical and citizen-reported data show shifts in bumblebee phenology over the last century in Sweden. Biodiversity and Conservation, 32(5), 1523–1547. https://doi.org/10.1007/s10531-023-02563-5
Prestele, R., Brown, C., Polce, C., Maes, J., & Whitehorn, P. (2021). Large variability in response to projected climate and land‐use changes among European bumblebee species. Global Change Biology, 27(19), 4530–4545. https://doi.org/10.1111/gcb.15780
Li, D., Barve, N., Brenskelle, L., Earl, K., Barve, V., Belitz, M. W., Doby, J., Hantak, M. M., Oswald, J. A., Stucky, B. J., Walters, M., & Guralnick, R. P. (2021). Climate, urbanization, and species traits interactively drive flowering duration. Global Change Biology, 27(4), 892–903. https://doi.org/10.1111/gcb.15461
iNaturalist community. Observations of Bombus spp. from Oregon, recorded between 2017 and 2024. Data exported from https://www.inaturalist.org on December 4, 2024.
